---
title: "01 Census Pull"
output: html_document
date: "2025-10-03"
---

# Check for packages and install them if not already installed

```{r setup, message=FALSE}
# List all packages needed for this notebook
required_packages <- c("httr", "jsonlite", "dplyr", "DBI", "RSQLite")

# Check if each one is installed, and install it if missing
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
# Confirm
cat("All packages loaded successfully!\n")
```

# Setup

# This notebook pulls Census data using the API and prepares it for later steps.

```{r load packages and set up test}
# Load required packages
library(httr)
library(jsonlite)
library(dplyr)


# Simple test request to Census API
url <- "https://api.census.gov/data/2020/dec/pl?get=P1_001N,NAME&for=state:*"
res <- GET(url)
census_raw <- fromJSON(content(res, "text"))

# Check first 5 rows
head(census_raw)
```

# Set up lists of variables for API Pull from Census

### Three pulls are required due to the number of variables and some being decennial specific

```{R Census variable setup}
# --- Census variable + URL setup (converted from your Python notebook) ---
# Base URLs
census_base_url <- "https://api.census.gov/"
census_dataset_url <- "data/2022/acs/acs5"
decennial_dataset_url <- "data/2020/dec/pl"

# Variables as vectors (easier to edit than a long single string)
census_variables1 <- c(
  "GEO_ID","NAME","B01001_001E","B25034_001E","B25034_002E","B25034_003E",
  "B25034_004E","B25034_005E","B25034_006E","B25034_007E","B25034_008E",
  "B25034_009E","B25034_010E","B25034_011E","B25046_001E","B27010_017E",
  "B27010_033E","B27010_050E","B27010_066E","B28002_013E","C16002_004E",
  "C16002_007E","C16002_010E","C16002_013E","B09019_005E","B09019_008E",
  "B19101_001E","C17002_002E","C17002_003E","C17002_004E","C17002_005E",
  "C17002_006E","C17002_007E","B20005_002E","B01001_003E","B01001_004E",
  "B01001_005E","B01001_006E","B01001_020E","B01001_021E"
)

census_variables2 <- c(
  "GEO_ID","C16002_004E","C16002_007E","C16002_010E","C16002_013E",
  "B05013_017E","B05013_018E","B05013_019E","B16002_022E","B16002_025E",
  "B01001_017E","B01001_018E","B01001_019E","B01001_016E","B16002_028E",
  "B16002_031E","B16002_034E","B16002_037E","B01001_044E","B01001_045E",
  "B01001_046E","B01001_047E","B01001_048E","B01001_049E","B02001_002E",
  "B02001_003E","B02001_004E","B02001_005E","B02001_006E","B02001_007E",
  "B02001_008E","B25008_003E"
)

decennial_variables3 <- c(
  "GEO_ID","P5_002N","P5_003N","P5_004N","P5_005N","P5_008N","P5_009N","P5_010N"
)
```

```{R combine variables and generate URLs}
# Collapse to comma-separated strings for the API 'get=' parameter
census_variables1_str <- paste(census_variables1, collapse = ",")
census_variables2_str <- paste(census_variables2, collapse = ",")
decennial_variables3_str <- paste(decennial_variables3, collapse = ",")

# Geography clause (URL-encoded spaces). State:15 = Hawaii
# Change "state:15" to a different state FIPS if needed
in_geography <- "&for=block%20group:*&in=state:15%20county:*%20tract:*"

# Combined endpoints
census_combined_url <- paste0(census_base_url, census_dataset_url)
decennial_combined_url <- paste0(census_base_url, decennial_dataset_url)

# Full query URLs (ready to GET)
cbg_url1 <- paste0(census_combined_url, "?get=", census_variables1_str, in_geography)
cbg_url2 <- paste0(census_combined_url, "?get=", census_variables2_str, in_geography)
cbg_url3 <- paste0(decennial_combined_url, "?get=", decennial_variables3_str, in_geography)

# Variable metadata (variables.json) URLs
census_fields_url1 <- paste0(census_combined_url, "/variables.json")
census_fields_url2 <- paste0(census_combined_url, "/subject/variables.json")
census_fields_url3 <- paste0(decennial_combined_url, "/variables.json")

# Quick preview (print the start of each URL to verify)
#cat("CBG URL 1 (start):\n", substr(cbg_url1, 1, 200), "...\n\n")
#cat("CBG URL 2 (start):\n", substr(cbg_url2, 1, 200), "...\n\n")
#cat("CBG URL 3 (start):\n", substr(cbg_url3, 1, 200), "...\n\n")


```

```{R Print full URLs for testing}
cat("CBG URL 1:\n", cbg_url1, "\n\n")
cat("CBG URL 2:\n", cbg_url2, "\n\n")
cat("CBG URL 3:\n", cbg_url3, "\n\n")
```

# Define FUNCTION to pull data JSON from Census API and store in lists

```{R Define fetch function}
# Define a small helper function to fetch and convert JSON to data.frame
fetch_census <- function(url) {
  resp <- httr::GET(url)
  if (resp$status_code != 200) {
    stop(paste("Request failed with status", resp$status_code))
  }
  
  # Convert JSON text to list, then to data.frame
  raw_json <- httr::content(resp, "text")
  data_list <- jsonlite::fromJSON(raw_json)
  
  # The first row of Census API results contains the column names
  df <- as.data.frame(data_list[-1, ], stringsAsFactors = FALSE)
  names(df) <- data_list[1, ]
  
  return(df)
}
```

```{R Run function to pull data from Census API}
# Fetch each dataset
cbg1 <- fetch_census(cbg_url1)
cbg2 <- fetch_census(cbg_url2)
cbg3 <- fetch_census(cbg_url3)

# Print quick summaries to verify
cat("✅ Downloads complete!\n")
cat("Rows in cbg1:", nrow(cbg1), "\n")
cat("Rows in cbg2:", nrow(cbg2), "\n")
cat("Rows in cbg3:", nrow(cbg3), "\n\n")

# Peek at the structure of one
head(cbg1[, 1:6])
```

## Merge first two dataframes into new one called cbg

```{r Merge Second DF with First, echo=TRUE}
# Drop duplicate columns (except GEO_ID) from cbg2
duplicate_columns <- intersect(names(cbg2), names(cbg1))
duplicate_columns <- setdiff(duplicate_columns, "GEO_ID")

if (length(duplicate_columns) > 0) {
  message("Dropping duplicate columns: ", paste(duplicate_columns, collapse = ", "))
  cbg2 <- cbg2[ , !(names(cbg2) %in% duplicate_columns)]
}

# Merge cbg1 and cbg2 on GEO_ID
cbg_df <- merge(cbg1, cbg2, by = "GEO_ID", all.x = TRUE, suffixes = c("", "_2"))

# Quick check
cat("Merged dataframe has", nrow(cbg_df), "rows and", ncol(cbg_df), "columns\n")
head(cbg_df)
```

## Merge third dataframe with existing one called cbg_df

```{R Merge Third DF with previously merged}
# Drop duplicate columns (except GEO_ID) from cbg3
duplicate_columns <- intersect(names(cbg3), names(cbg_df))
duplicate_columns <- setdiff(duplicate_columns, "GEO_ID")

if (length(duplicate_columns) > 0) {
  message("Dropping duplicate columns: ", paste(duplicate_columns, collapse = ", "))
  cbg3 <- cbg3[ , !(names(cbg3) %in% duplicate_columns)]
}

# Merge cbg (from previous step) with cbg3
cbg_df <- merge(cbg_df, cbg3, by = "GEO_ID", all.x = TRUE, suffixes = c("", "_3"))

# Quick check
cat("Merged dataframe now has", nrow(cbg_df), "rows and", ncol(cbg_df), "columns\n")
head(cbg_df)
```

## Replace nulls in the dataframe with zeroes

```{R replace nulls with zeroes}
cbg_df[is.na(cbg_df)] <- 0

# Check if there are any nulls left
print(any(is.na(cbg_df)))

```

```{R convert numbers in DF to actual numeric fields}
# Loop over all columns
for (colname in names(cbg_df)) {
  # Try converting to numeric
  numeric_col <- suppressWarnings(as.numeric(cbg_df[[colname]]))
  
  # If there are no new NAs introduced, replace the column
  if (!all(is.na(numeric_col) & !is.na(cbg_df[[colname]]))) {
    cbg_df[[colname]] <- numeric_col
  }
}

# Quick check: see the classes of all columns
sapply(cbg_df, class)
```

## Create and calculate field for Limited English Speaking Households

```{R create and calculate LimitedEnglishH field}

# Check input fields first
print(sapply(cbg_df[, c("C16002_004E", "C16002_007E", "C16002_010E", "C16002_013E")], class))
print(head(cbg_df[, c("C16002_004E", "C16002_007E", "C16002_010E", "C16002_013E")], 10))

# Create a new column by summing other columns
cbg_df$LimitedEnglishHH <- cbg_df$C16002_004E + 
                        cbg_df$C16002_007E + 
                        cbg_df$C16002_010E + 
                        cbg_df$C16002_013E

# Preview the first few rows
head(cbg_df$LimitedEnglishHH)
```

## Create \output folder

```{R create output folder}
output_folder_path <- file.path(getwd(), "output")

# Print it to confirm
cat("Output folder:", output_folder_path, "\n")

if (!dir.exists(output_folder_path)) {
  dir.create(output_folder_path)
  cat("Created output folder.\n")
}

```

```{R generate names for CSV files}
# Define output file names
cbg_table_name <- "census_data"

# (Optional) show full file paths
cbg_csv_path <- file.path(output_folder_path, paste0(cbg_table_name, ".csv"))


cat("CBG CSV path:", cbg_csv_path, "\n")
```

```{R save CSV files for data}
# Build full CSV file path
cbg_csv_path <- file.path(output_folder_path, paste0(cbg_table_name, ".csv"))

# Write dataframe to CSV
write.csv(cbg_df, cbg_csv_path, row.names = FALSE, fileEncoding = "UTF-8")

cat("CSV file created successfully:", cbg_csv_path, "\n")
```

``` {R save Census data to SQLite database}
# Define the database path
sqlite_path <- file.path(output_folder_path, "census_data.sqlite")

# Connect to (or create) the SQLite database
conn <- dbConnect(RSQLite::SQLite(), sqlite_path)

# Write the table (overwrite if it exists)
dbWriteTable(conn, "census_data", cbg_df, overwrite = TRUE)

# Optionally list tables to confirm
cat("Tables in database:\n")
print(dbListTables(conn))

# Disconnect
dbDisconnect(conn)

cat("SQLite database created successfully:", sqlite_path, "\n")
```